---
title: "New York City Shooting Incidents"
author: "Seth Porter"
date: "2023-03-20"
output:
  powerpoint_presentation:
    reference_doc: custom-reference.pptx
  beamer_presentation:
    theme: "Pittsburgh"
    colortheme: "dolphin"
    fonttheme: "structurebold"
---

```{r setup, include=FALSE}
library(knitr)
knitr::opts_chunk$set(echo = FALSE)
```

```{r setup_data, echo=FALSE, include=FALSE}
library(conflicted)

# Tidyverse and sub-projects
library(tidyverse)
# if this fails, please install with `install.packages("tidyverse")` in the console
library(lubridate)

# The new fancy error messages breaks LaTeX rendering at least on
# my machine, so avoid it.
conflict_prefer("filter", "dplyr")
conflict_prefer("lag", "dplyr")

url <- "https://data.cityofnewyork.us/api/views/833y-fsy8/rows.csv?accessType=DOWNLOAD"
raw_incidents <- read_csv(url)
gender_levels <- c("M", "F", "U")
# Use the data to generate these lists, the values are long and messy
race_levels <- levels(factor(raw_incidents$VIC_RACE))
# PERP_AGE_GROUP has some values that VIC_AGE_GROUP doesn't, but they're bad.
age_levels = levels(factor(raw_incidents$VIC_AGE_GROUP))
boro_levels <- c( "BRONX", "BROOKLYN", "MANHATTAN", "QUEENS", "STATEN ISLAND")
# use parse_factor to warn if we get unexpected values
incidents <- raw_incidents %>%
    filter(!is.na(JURISDICTION_CODE)) %>%
    mutate(OCCUR_DATE=mdy(OCCUR_DATE),
           LOCATION_DESC=replace_na(LOCATION_DESC, "NONE"),
           PERP_RACE=replace_na(PERP_RACE, "UNKNOWN"),
           PERP_AGE_GROUP=replace_na(PERP_AGE_GROUP, "UNKNOWN"),
           PERP_SEX=replace_na(PERP_SEX, "U")) %>%
    mutate(PERP_AGE_GROUP=recode(PERP_AGE_GROUP,
           "1020"= "UNKNOWN",
           "224" = "UNKNOWN",
           "940" = "UNKNOWN")) %>%
    mutate(PERP_SEX=parse_factor(PERP_SEX, levels=gender_levels),
           VIC_SEX=parse_factor(VIC_SEX, levels=gender_levels),
           PERP_RACE=parse_factor(PERP_RACE, levels=race_levels),
           VIC_RACE=parse_factor(VIC_RACE, levels=race_levels),
           BORO=parse_factor(BORO, levels=boro_levels),
           # PRECINCT=factor(PRECINCT),
           LOCATION_DESC=factor(LOCATION_DESC),
           JURISDICTION_CODE = fct_recode(factor(JURISDICTION_CODE),
                                          "Patrol" = "0",
                                          "Transit" = "1",
                                          "Housing" = "2"),
           PERP_AGE_GROUP=parse_factor(PERP_AGE_GROUP, age_levels),
           VIC_AGE_GROUP=parse_factor(VIC_AGE_GROUP, age_levels)) %>%
    mutate(year=floor_date(OCCUR_DATE, "year")) %>%
    select(-c("Lon_Lat", "INCIDENT_KEY", "X_COORD_CD", "Y_COORD_CD", "Latitude", "Longitude", "PRECINCT"))
```

# What are we looking at, and why should I care?

We will examine "shooting incident" data from the New York City Police Department.

Shooting incidents are extreme events which can end or radically change the
lives of both the victim and the shooter in a single action.

Understanding their patterns is crucial to policy and intervention decisions
from law enforcement, other city agencies, and non-governmental actors.

# Where is the data from?

* Dataset sourced from <https://catalog.data.gov/dataset/nypd-shooting-incident-data-historic>, 
  described in Footnotes (<https://bit.ly/3KSLRjA>)
* "List of every shooting incident that occurred in NYC going back to 2006 through the end of the previous calendar year."
* "Only valid shooting incidents resulting in an injured victim are included in this
release."

# Secondary data sources

In addition to the primary data, some secondary sources are used to
contextualize or suggest possible causal variables:

* Temperature data from <https://www.weather.gov/media/okx/Climate/CentralPark/DailyAvgTNormals.pdf>
* Population data from <https://www.census.gov/quickfacts/newyorkcitynewyork>

# Preliminary Cautions

Be cautious; this data is from a law enforcement agency reporting on its own
jurisdiction, and may have obvious
as well as unexpected biases or gaps in the collection and reporting process.

# Self Links

Github repo: <https://github.com/symmatree/data-science/tree/main/r>

* These slides: Source (<https://github.com/symmatree/data-science/raw/main/r/NYPDShootingDataSlides.Rmd>) /
  [pdf](https://github.com/symmatree/data-science/raw/main/r/NYPDShootingDataSlides.pdf) /
  [pptx](https://github.com/symmatree/data-science/raw/main/r/NYPDShootingDataSlides.pptx)
* The full analysis document has more discussion of data tidying and methods:
  <https://github.com/symmatree/data-science/raw/main/r/NYPDShootingData.Rmd> (source) /
  [pdf](https://github.com/symmatree/data-science/raw/main/r/NYPDShootingData.pdf)

# Two Major Approaches

We will consider

* patterns-over-time and seasonality
* demographics of victims and perpetrators

in hopes of understanding the patterns and impacts of these incidents.

# Data Analysis - Temporal trends

Our first layer of analysis will focus on **when** shootings occur, considering
both the overall trend and various scales of periodic behavior.

# Long-term trends

::: columns

:::: column

Reported shooting incidents dropped steadily, then
rose sharply in 2020 (presumably pandemic-related, though possibly
a reporting change).

This is such a sharp reversal that it dominates other trends.

The rest of this analysis will normalize by annual incident count to 
reveal patterns masked by this macro-trend.

::::

:::: column

```{r hist_DATE_year, echo=FALSE, dpi=300}
yearly <- incidents %>% count(
  year=floor_date(OCCUR_DATE, "year"),
  year_num=year(floor_date(OCCUR_DATE, "year")))
yearly_fit <- lm(yearly$n ~ yearly$year)

yearly_with_pred <- yearly %>%
  mutate(predicted=predict(yearly_fit))

yearly_with_pred %>%
  ggplot(aes(x=year)) +
    ggtitle("Shooting Incidents over time") +
    theme(legend.position="bottom") +
    ggplot2::guides(colour=ggplot2::guide_legend(title="")) +
    ylab("Number of incidents") +
    geom_point(aes(y=n, color="actual")) +
    geom_line(aes(y=n, color="actual")) +
    geom_line(aes(y=predicted, color="predicted"))
```
::::

:::

# Seasonality: Month of year

::: columns

:::: column

A shifted and vertically-scaled sine wave fits the monthly data quite nicely.

Average NYC temperatures correlate strongly with both this model and the
observed incident counts.

This suggests (though does not prove!) that temperature-driven behavior patterns
are a major driver of the overall level of shootings.

::::

:::: column

```{r plot_monthly_sine_model, dpi=300}
monthly = incidents %>%
  count(month=month(floor_date(OCCUR_DATE, "month"), label=TRUE),
        month_num=month(floor_date(OCCUR_DATE, "month")),
        year=floor_date(OCCUR_DATE, "year")) %>%
  left_join(yearly, by="year", suffix=c("", ".total")) %>%
  mutate(fraction=n/n.total)

sine_model <- nls(fraction ~ v_scale*sin(2*pi*month_num/12.0+phase)+offset, data=monthly, start=list(v_scale=1, phase=7*2*pi/12, offset=0.1))

month_df = tibble(month_num=seq(1, 12))
month_df$predicted=predict(sine_model, newdata=month_df)

# Per https://www.weather.gov/media/okx/Climate/CentralPark/DailyAvgTNormals.pdf
nyc_average_temps = c(33.6, 35.9, 43.0, 53.7, 63.2, 72.1, 77.6, 76.1, 69.3, 58, 48.1, 39.1)
month_df$average_temp <- nyc_average_temps

# Let the computer figure out the right scale and offset to align the temp
# data with the sine model outputs.
temp_model <- nls(predicted ~ temp_scale*average_temp+temp_offset,
                  data=month_df,
                  start=list(temp_scale=1, temp_offset=0))
month_df$scaled_temp <- predict(temp_model)


ggplot() +
    ggtitle("Shooting Incidents by Month") +
    theme(legend.position="bottom") +
    ggplot2::guides(linetype=ggplot2::guide_legend(title=""), 
                    color=ggplot2::guide_colourbar(title="")) +
    ylab("Fraction of annual incidents") +
    geom_point(data=monthly, mapping=aes(x=month, y=fraction, color=year_num)) +
    geom_line(data=month_df, mapping=aes(x=month_num, y=predicted, linetype="Sine Model")) +
    geom_line(data=month_df, mapping=aes(x=month_num, y=scaled_temp, linetype="Avg Temperature"))

```
::::

:::


# Seasonality: Day of week

::: columns

:::: column

Weekly rhythms have a strong influence on shooting incidents; a scaled and
phase-shifted sine wave fits quite nicely (RMSE \~2 percentage points of
annual incidents).

There is no immediately-obvious explanation of this pattern, other than the
incident rates being higher on Fridays and weekends when more social interaction
is likely to occur.

::::

:::: column

```{r by_day_of_week, dpi=300}
daily = incidents %>%
  count(day_of_week=wday(OCCUR_DATE, label=TRUE),
        day_num=wday(OCCUR_DATE),
        year=floor_date(OCCUR_DATE, "year")) %>%
  left_join(yearly, by="year", suffix=c("", ".total")) %>%
  # n.total is the total incidents in the given year
  mutate(fraction=n/n.total)

sine_day_model <- nls(fraction ~ v_scale*sin(2*pi*day_num/7.0+phase)+offset,
                      data=daily, start=list(v_scale=1, phase=0, offset=0.1))

day_pred = tibble(day_num=seq(1, 7))
day_pred$predicted=predict(sine_day_model, newdata=day_pred)
day_pred$day_of_week = levels(daily$day_of_week)

ggplot() +
  ggtitle("Shooting Incidents by Day of week") +
  theme(legend.position="bottom") +
  ggplot2::guides(shape=ggplot2::guide_legend(title="")) +
  ylab("Fraction of annual incidents") +
  geom_point(data=daily, mapping=aes(x=day_of_week, y=fraction, color=year, shape="actual")) +
  geom_point(data=day_pred, mapping=aes(x=day_num, y=predicted, shape="Sine model", size=2)) +
  geom_line(data=day_pred, mapping=aes(x=day_num, y=predicted)) +
  guides(size="none")
```

::::

:::

# Seasonality: Hour of day

::: columns

:::: column

While not a simple sinusoidal, there is a repeated cycle
throughout the day, with incidents rarest in the morning hours
and most common at night until 4am (which is when bars close in NYC).

This aligns with the weekly pattern: shootings happen less
during typical working hours and days.

::::

:::: column

```{r hour_of_day, dpi=300}
hour_of_day = incidents %>%
  count(hour_of_day=hour(OCCUR_TIME),
        year=floor_date(OCCUR_DATE, "year")) %>%
  left_join(yearly, by="year", suffix=c("", ".total")) %>%
  # n.total is the total incidents in the given year
  mutate(fraction=n/n.total)

hour_of_day %>%
  ggplot(aes(hour_of_day, fraction, color=year)) +
    ggtitle("Shooting Incidents by hour of day") +
    theme(legend.position="bottom") +
    ylab("Fraction of annual incidents") +
    geom_point()

```
::::

:::


# Demographics: Victim, By Race

::: columns

:::: column

The victim's race is overwhelmingly
coded as Black.

::::

:::: column

```{r vic_race, echo=FALSE, dpi=300}
incidents %>% 
  count(year=year, VIC_RACE=VIC_RACE) %>%
  left_join(incidents %>% count(year=year),
            by="year", suffix=c("", ".total")) %>%
  mutate(fraction=n/n.total) %>%
  ggplot(aes(year, fraction, color=VIC_RACE)) +
    ggtitle("Victim Demographics by Race") +
    ggplot2::theme(
      legend.text=ggplot2::element_text(size=8),
      legend.box.margin = margin(6, 6, 6, 6),
      legend.position="bottom") +
    ggplot2::guides(color=ggplot2::guide_legend(ncol=2, title="")) +
    ylab("Fraction of annual incidents") +
    geom_line() +
    geom_point()
```
::::

:::

# Demographics: Perpetrator, by Race


::: columns

:::: column

The same holds true, in almost the same proportions, for
the perpetrator's race (where known).

::::

:::: column

```{r perp_race, echo=FALSE, dpi=300}
known_perp <- incidents %>%
    filter((PERP_RACE != "UNKNOWN") & (PERP_AGE_GROUP != "UNKNOWN"))
known_perp %>% 
  count(year=year, PERP_RACE=PERP_RACE) %>%
  left_join(known_perp %>% count(year=year),
            by="year", suffix=c("", ".total")) %>%
  mutate(fraction=n/n.total) %>%
  ggplot(aes(year, fraction, color=PERP_RACE)) +
    ggtitle("Perpetrator Demographics by Race, where known") +
    ggplot2::theme(
      legend.text=ggplot2::element_text(size=8),
      legend.box.margin = margin(6, 6, 6, 6),
      legend.position="bottom") +
    ggplot2::guides(color=ggplot2::guide_legend(ncol=2, title="")) +
    ylab("Fraction of annual incidents") +
    geom_line() +
    geom_point()
```

::::

:::

# Younger Black males

::: columns

:::: column

We can narrow the data to consider only the combined fractions for victims and
perpetrators coded as Black males in the 18-24 or 25-44 age groups.

More than half of all shooting incidents involve someone in this demographic
as victim, and roughly the same proportion as perpetrator (where known).

The entire Black population of NYC is just under a quarter of the city; males
in this age group would be a still smaller fraction.

::::

:::: column

```{r aggregate_demo, echo=FALSE, dpi=300}
known_perp %>%
  count(year, PERP_RACE, PERP_SEX, PERP_AGE_GROUP) %>%
  left_join(known_perp %>% count(year=year),
            by="year", suffix=c("", ".total")) %>%
  mutate(fraction=n/n.total) %>%
  filter(fraction > 0.15) %>%
  group_by(year) %>%
  summarize(perp_fraction=sum(fraction)) %>%
  left_join(incidents %>%
    count(year, VIC_RACE, VIC_SEX, VIC_AGE_GROUP) %>%
    left_join(incidents %>% count(year=year),
              by="year", suffix=c("", ".total")) %>%
    mutate(fraction=n/n.total) %>%
    filter(fraction > 0.15) %>%
    group_by(year) %>%
    summarize(vic_fraction=sum(fraction)),
    by="year") %>%
  ggplot(aes(x=year)) +
    ggtitle("Shooting Incidents, Black males age 18-44") +
    ggplot2::theme(
      legend.text=ggplot2::element_text(size=8),
      legend.box.margin = margin(6, 6, 6, 6),
      legend.position="bottom") +
    ggplot2::guides(color=ggplot2::guide_legend(ncol=1, title="")) +
    ylim(0, 1) +
    ylab("Fraction of Annual Shootings") +
    geom_point(aes(y=perp_fraction, color="Perpetrator (where known)")) +
    geom_line(aes(y=perp_fraction, color="Perpetrator (where known)")) +
    geom_point(aes(y=vic_fraction, color="Victim")) +
    geom_line(aes(y=vic_fraction, color="Victim")) +
    geom_hline(aes(yintercept=0.238, color="Black Population"), linetype=2)
```

::::

:::

# Conclusions - Temporal Conclusions

Despite sharp changes in overall incident counts, there are strong seasonal
patterns at multiple scales: over the course of a day, a week, and a year

# Conclusions - Demographic Conclusions

The impact of shootings (both as victim and perpetrator) falls wildly
disproportionately on individuals reported as Black males aged 18-44.

Any strategy or intervention must consider both these temporal and 
demographic factors or it will be mis-targeted or inefficient.

# A Caution

The demographic concentration of shootings should not be read as *causal*
but rather as descriptive and correlated. Demographics are correlated with
many other factors including socio-economic status, education, job opportunities,
geographic location, historical deprivation etc., which are more likely to
be actual causation.

# Error and Bias

Finally, we consider the many ways this analysis might be wrong or misleading.

# Error and Bias: Data Source Motivations

The underlying data comes from the NYPD. They are not a neutral party:
this data may be used to demonstrate their effectiveness, to argue for
more resources in certain areas, or to support theories of crime and
victimhood.

# Error and Bias: Data Source Avenues of Bias

Any of these would be a reason to bias the data collection and reporting:

* Skewing demographic data toward desirable patterns
* More subtly, controlling what is reported as a shooting incident and what is
  absent altogether

# Error and Bias: Analysis

Beyond the data, the analysis may be flawed, biased or misleading.

* The focus on proportional analysis masks the changing absolute numbers of
  incidents, which may be a better measure of "cost" in many dimensions
* There may be important subtleties in who reports data and how, unknown to
  the analyst. Demographic fields are subjective and vulnerable to misinterpretation.

That said, the patterns reported *seem* to be so strong as to resist most
small-scale biases or misinterpretations!