---
title: "New York City Shooting Incidents"
author: "Seth Porter"
date: "2023-03-20"
output:
  beamer_presentation:
    theme: "Pittsburgh"
    colortheme: "dolphin"
    fonttheme: "structurebold"
  powerpoint_presentation:
    reference_doc: custom-reference.pptx
---

```{r setup, include=FALSE}
library(knitr)
knitr::opts_chunk$set(echo = FALSE)
```

```{r setup_data, echo=FALSE, include=FALSE}
library(conflicted)

# Tidyverse and sub-projects
library(tidyverse)
# if this fails, please install with `install.packages("tidyverse")` in the console
library(lubridate)

# The new fancy error messages breaks LaTeX rendering at least on
# my machine, so avoid it.
conflict_prefer("filter", "dplyr")
conflict_prefer("lag", "dplyr")

url <- "https://data.cityofnewyork.us/api/views/833y-fsy8/rows.csv?accessType=DOWNLOAD"
raw_incidents <- read_csv(url)
gender_levels <- c("M", "F", "U")
# Use the data to generate these lists, the values are long and messy
race_levels <- levels(factor(raw_incidents$VIC_RACE))
# PERP_AGE_GROUP has some values that VIC_AGE_GROUP doesn't, but they're bad.
age_levels = levels(factor(raw_incidents$VIC_AGE_GROUP))
boro_levels <- c( "BRONX", "BROOKLYN", "MANHATTAN", "QUEENS", "STATEN ISLAND")
# use parse_factor to warn if we get unexpected values
incidents <- raw_incidents %>%
    filter(!is.na(JURISDICTION_CODE)) %>%
    mutate(OCCUR_DATE=mdy(OCCUR_DATE),
           LOCATION_DESC=replace_na(LOCATION_DESC, "NONE"),
           PERP_RACE=replace_na(PERP_RACE, "UNKNOWN"),
           PERP_AGE_GROUP=replace_na(PERP_AGE_GROUP, "UNKNOWN"),
           PERP_SEX=replace_na(PERP_SEX, "U")) %>%
    mutate(PERP_AGE_GROUP=recode(PERP_AGE_GROUP,
           "1020"= "UNKNOWN",
           "224" = "UNKNOWN",
           "940" = "UNKNOWN")) %>%
    mutate(PERP_SEX=parse_factor(PERP_SEX, levels=gender_levels),
           VIC_SEX=parse_factor(VIC_SEX, levels=gender_levels),
           PERP_RACE=parse_factor(PERP_RACE, levels=race_levels),
           VIC_RACE=parse_factor(VIC_RACE, levels=race_levels),
           BORO=parse_factor(BORO, levels=boro_levels),
           # PRECINCT=factor(PRECINCT),
           LOCATION_DESC=factor(LOCATION_DESC),
           JURISDICTION_CODE = fct_recode(factor(JURISDICTION_CODE),
                                          "Patrol" = "0",
                                          "Transit" = "1",
                                          "Housing" = "2"),
           PERP_AGE_GROUP=parse_factor(PERP_AGE_GROUP, age_levels),
           VIC_AGE_GROUP=parse_factor(VIC_AGE_GROUP, age_levels)) %>%
    mutate(year=floor_date(OCCUR_DATE, "year")) %>%
    select(-c("Lon_Lat", "INCIDENT_KEY", "X_COORD_CD", "Y_COORD_CD", "Latitude", "Longitude", "PRECINCT"))
```

# What are we looking at, and why should I care?

We will examine "shooting incident" data from the New York City Police Department.

Shooting incidents are extreme events which can end or radically change the
lives of both the victim and the shooter in a single action.

Understanding their patterns is crucial to policy and intervention decisions
from law enforcement, other city agencies, and non-governmental actors.

::: notes

What are we looking at? Why should you care?

* Shootings end or radically change the lives of both the victim and the shooter.
* Understanding the patterns to shootings is vital when trying to reduce them
  or mitigate their effects.

So this is data about a subject that's important to people at both a personal
and a systemic level.

:::

# Where is the data from?

* Dataset sourced from <https://catalog.data.gov/dataset/nypd-shooting-incident-data-historic>, 
  described in Footnotes (<https://bit.ly/3KSLRjA>)
* "List of every shooting incident that occurred in NYC going back to 2006 through the end of the previous calendar year."
* "Only valid shooting incidents resulting in an injured victim are included in this
release."

::: notes
This is a dataset from the NYPD, with links here to the data and a PDF describing it.

This is a list of every shooting incident that occurred in New York since 2006, where someone was shot and injured.
(And, it should be pointed out, where that was reported to or discovered by police.)
:::

# Secondary data sources

In addition to the primary data, some secondary sources are used to
contextualize or suggest possible causal variables:

* Temperature data from <https://www.weather.gov/media/okx/Climate/CentralPark/DailyAvgTNormals.pdf>
* Population data from <https://www.census.gov/quickfacts/newyorkcitynewyork>

::: notes
I used some additional data from public sources for temperature and population.
:::

# Preliminary Cautions

Be cautious; this data is from a agency reporting on its own
jurisdiction, and may have resulting biases or gaps in the collection,
characterization, and reporting process.

::: notes
Before presenting the data, I just want to remind everyone that this is from
an organization reporting on its own area of responsibility, which is an
inherent conflict of interest (or rather a family of related conflicts,
at different levels and in different parts of the organization).
:::

# Self Links

Github repo: <https://github.com/symmatree/data-science/tree/main/r>

* These slides: Source (<https://github.com/symmatree/data-science/raw/main/r/NYPDShootingDataSlides.Rmd>) /
  [pdf](https://github.com/symmatree/data-science/raw/main/r/NYPDShootingDataSlides.pdf) /
  [pptx](https://github.com/symmatree/data-science/raw/main/r/NYPDShootingDataSlides.pptx)
* The full analysis document has more discussion of data tidying and methods:
  <https://github.com/symmatree/data-science/raw/main/r/NYPDShootingData.Rmd> (source) /
  [pdf](https://github.com/symmatree/data-science/raw/main/r/NYPDShootingData.pdf)

::: notes
These slides are available on Github, as well as the full analysis
which gets further into many of these topics including data cleaning.
:::

# Two Major Approaches

We will consider

* patterns-over-time and seasonality
* demographics of victims and perpetrators

in hopes of understanding the patterns and impacts of these incidents.

::: notes
In this analysis I consider the patterns of these incidents in terms of
1) time and 2) the demographics of the victims and perpetrators.
:::

# Data Analysis - Temporal trends

Our first layer of analysis will focus on **when** shootings occur, considering
both the overall trend and various scales of periodic behavior.

::: notes
We look first at the time axis, on several different scales.
:::

# Long-term trends

::: columns

:::: column

Reported shooting incidents dropped steadily, then
rose sharply in 2020 (presumably pandemic-related, though possibly
a reporting change).

This is such a sharp reversal that it dominates other trends.

The rest of this analysis will normalize by annual incident count to 
reveal patterns masked by this macro-trend.

::::

:::: column

```{r hist_DATE_year, echo=FALSE, dpi=300}
yearly <- incidents %>% count(
  year=floor_date(OCCUR_DATE, "year"),
  year_num=year(floor_date(OCCUR_DATE, "year")))
yearly_fit <- lm(yearly$n ~ yearly$year)

yearly_with_pred <- yearly %>%
  mutate(predicted=predict(yearly_fit))

yearly_with_pred %>%
  ggplot(aes(x=year)) +
    ggtitle("Shooting Incidents over time") +
    theme(legend.position="bottom") +
    ggplot2::guides(colour=ggplot2::guide_legend(title="")) +
    ylab("Number of incidents") +
    geom_point(aes(y=n, color="actual")) +
    geom_line(aes(y=n, color="actual")) +
    geom_line(aes(y=predicted, color="predicted"))
```
::::

:::

::: notes
As you see in the chart, an initial roughly-linear trend through 2019 turned
into a steep rise in 2020, sustained in 2021.
This two-stage macro behavior dominates everything else. 
Therefore, for the rest of this presentation, all graphs will be normalized
to “fraction of the incidents that year” to reveal patterns otherwise masked.
:::


# Seasonality: Month of year

::: columns

:::: column

A shifted and vertically-scaled sine wave fits the monthly data quite nicely.

Average NYC temperatures correlate strongly with both this model and the
observed incident counts.

This suggests (though does not prove!) that temperature-driven behavior patterns
are a major driver of the overall level of shootings.

::::

:::: column

```{r plot_monthly_sine_model, dpi=300}
monthly = incidents %>%
  count(month=month(floor_date(OCCUR_DATE, "month"), label=TRUE),
        month_num=month(floor_date(OCCUR_DATE, "month")),
        year=floor_date(OCCUR_DATE, "year")) %>%
  left_join(yearly, by="year", suffix=c("", ".total")) %>%
  mutate(fraction=n/n.total)

sine_model <- nls(fraction ~ v_scale*sin(2*pi*month_num/12.0+phase)+offset, data=monthly, start=list(v_scale=1, phase=7*2*pi/12, offset=0.1))

month_df = tibble(month_num=seq(1, 12))
month_df$predicted=predict(sine_model, newdata=month_df)

# Per https://www.weather.gov/media/okx/Climate/CentralPark/DailyAvgTNormals.pdf
nyc_average_temps = c(33.6, 35.9, 43.0, 53.7, 63.2, 72.1, 77.6, 76.1, 69.3, 58, 48.1, 39.1)
month_df$average_temp <- nyc_average_temps

# Let the computer figure out the right scale and offset to align the temp
# data with the sine model outputs.
temp_model <- nls(predicted ~ temp_scale*average_temp+temp_offset,
                  data=month_df,
                  start=list(temp_scale=1, temp_offset=0))
month_df$scaled_temp <- predict(temp_model)


ggplot() +
    ggtitle("Shooting Incidents by Month") +
    theme(legend.position="bottom") +
    ggplot2::guides(linetype=ggplot2::guide_legend(title=""), 
                    color=ggplot2::guide_colourbar(title="")) +
    ylab("Fraction of annual incidents") +
    geom_point(data=monthly, mapping=aes(x=month, y=fraction, color=year_num)) +
    geom_line(data=month_df, mapping=aes(x=month_num, y=predicted, linetype="Sine Model")) +
    geom_line(data=month_df, mapping=aes(x=month_num, y=scaled_temp, linetype="Avg Temperature"))

```
::::

:::

::: notes
Turning to variation within a year, we see a strong monthly pattern.
This can be approximated quite well by a scaled-and-shifted sine wave,
and more suggestively, by average temperature data for the city.

This suggests, though in no way proves, a connection between temperature,
behavior, and propensity for shootings.

Note that the fit is in no way perfect; that overshoot in February and
undershoot in December are notable.
:::

# Seasonality: Day of week

::: columns

:::: column

Weekly rhythms have a strong influence on shooting incidents; a scaled and
phase-shifted sine wave fits quite nicely (RMSE \~2 percentage points of
annual incidents).

There is no immediately-obvious explanation of this pattern, other than the
incident rates being higher on Fridays and weekends when more social interaction
is likely to occur.

::::

:::: column

```{r by_day_of_week, dpi=300}
daily = incidents %>%
  count(day_of_week=wday(OCCUR_DATE, label=TRUE),
        day_num=wday(OCCUR_DATE),
        year=floor_date(OCCUR_DATE, "year")) %>%
  left_join(yearly, by="year", suffix=c("", ".total")) %>%
  # n.total is the total incidents in the given year
  mutate(fraction=n/n.total)

sine_day_model <- nls(fraction ~ v_scale*sin(2*pi*day_num/7.0+phase)+offset,
                      data=daily, start=list(v_scale=1, phase=0, offset=0.1))

day_pred = tibble(day_num=seq(1, 7))
day_pred$predicted=predict(sine_day_model, newdata=day_pred)
day_pred$day_of_week = levels(daily$day_of_week)

ggplot() +
  ggtitle("Shooting Incidents by Day of week") +
  theme(legend.position="bottom") +
  ggplot2::guides(shape=ggplot2::guide_legend(title="")) +
  ylab("Fraction of annual incidents") +
  geom_point(data=daily, mapping=aes(x=day_of_week, y=fraction, color=year, shape="actual")) +
  geom_point(data=day_pred, mapping=aes(x=day_num, y=predicted, shape="Sine model", size=2)) +
  geom_line(data=day_pred, mapping=aes(x=day_num, y=predicted)) +
  guides(size="none")
```

::::

:::

::: notes
We see a strong weekly cycle as well, again fit nicely by a scaled-and-shifted
sine wave.

Weeks are a purely human phenomena, so the temperature explanation isn't
relevant here. This pattern is sustained across many years, so there must be
*some* underlying mechanism.
:::

# Seasonality: Hour of day

::: columns

:::: column

While not a simple sinusoidal, there is a repeated cycle
throughout the day, with incidents rarest in the morning hours
and most common at night until 4am (which is when bars close in NYC).

This aligns with the weekly pattern: shootings happen less
during typical working hours and days.

::::

:::: column

```{r hour_of_day, dpi=300}
hour_of_day = incidents %>%
  count(hour_of_day=hour(OCCUR_TIME),
        year=floor_date(OCCUR_DATE, "year")) %>%
  left_join(yearly, by="year", suffix=c("", ".total")) %>%
  # n.total is the total incidents in the given year
  mutate(fraction=n/n.total)

hour_of_day %>%
  ggplot(aes(hour_of_day, fraction, color=year)) +
    ggtitle("Shooting Incidents by hour of day") +
    theme(legend.position="bottom") +
    ylab("Fraction of annual incidents") +
    geom_point()

```
::::

:::

::: notes
The times of day when shootings occur are also highly structured. Not a sine
wave, but a tightly repeating pattern across the years.

One feature of this chart can likely be explained by the fact that bars close
at 4am in New York.
:::

# Conclusions - Temporal Conclusions

Despite sharp changes in overall incident counts, there are strong seasonal
patterns at multiple scales: over the course of a day, a week, and a year

# Demographics: Victim, By Race

::: columns

:::: column

The victim's race is overwhelmingly
coded as Black, disproportionate to the Black
share of the overall city population.

::::

:::: column

```{r vic_race, echo=FALSE, dpi=300}
incidents %>% 
  count(year=year, VIC_RACE=VIC_RACE) %>%
  left_join(incidents %>% count(year=year),
            by="year", suffix=c("", ".total")) %>%
  mutate(fraction=n/n.total) %>%
  ggplot(aes(year, fraction, color=VIC_RACE)) +
    ggtitle("Victim Demographics by Race") +
    ggplot2::theme(
      legend.text=ggplot2::element_text(size=8),
      legend.box.margin = margin(6, 6, 6, 6),
      legend.position="bottom") +
    ggplot2::guides(color=ggplot2::guide_legend(ncol=2, title="")) +
    ylab("Fraction of annual incidents") +
    geom_line() +
    geom_point() +
    geom_hline(aes(yintercept=0.238, color="Black Population (fraction of city)"), linetype=2)
```
::::

:::


::: notes
Turning to the demographics of the people involved, we start with the race of
the victim.
:::

# Demographics: Perpetrator, by Race


::: columns

:::: column

The same holds true, in almost the same proportions, for
the perpetrator's race (where known), again disproportionately so.

::::

:::: column

```{r perp_race, echo=FALSE, dpi=300}
known_perp <- incidents %>%
    filter((PERP_RACE != "UNKNOWN") & (PERP_AGE_GROUP != "UNKNOWN"))
known_perp %>% 
  count(year=year, PERP_RACE=PERP_RACE) %>%
  left_join(known_perp %>% count(year=year),
            by="year", suffix=c("", ".total")) %>%
  mutate(fraction=n/n.total) %>%
  ggplot(aes(year, fraction, color=PERP_RACE)) +
    ggtitle("Perpetrator Demographics by Race, where known") +
    ggplot2::theme(
      legend.text=ggplot2::element_text(size=8),
      legend.box.margin = margin(6, 6, 6, 6),
      legend.position="bottom") +
    ggplot2::guides(color=ggplot2::guide_legend(ncol=2, title="")) +
    ylab("Fraction of annual incidents") +
    geom_line() +
    geom_point() +
    geom_hline(aes(yintercept=0.238, color="Black Population (fraction of city)"), linetype=2)
```

::::

:::

# Younger Black males

::: columns

:::: column

We can narrow the data to consider only the combined fractions for victims and
perpetrators coded as Black males in the 18-24 or 25-44 age groups.

More than half of all shooting incidents involve someone in this demographic
as victim, and roughly the same proportion as perpetrator (where known).

The *entire* Black population of NYC is just under a quarter of the city, so
the corresponding line is an overestimate of the fraction of Black males in 
this age group.

::::

:::: column

```{r aggregate_demo, echo=FALSE, dpi=300}
known_perp %>%
  count(year, PERP_RACE, PERP_SEX, PERP_AGE_GROUP) %>%
  left_join(known_perp %>% count(year=year),
            by="year", suffix=c("", ".total")) %>%
  mutate(fraction=n/n.total) %>%
  filter(fraction > 0.15) %>%
  group_by(year) %>%
  summarize(perp_fraction=sum(fraction)) %>%
  left_join(incidents %>%
    count(year, VIC_RACE, VIC_SEX, VIC_AGE_GROUP) %>%
    left_join(incidents %>% count(year=year),
              by="year", suffix=c("", ".total")) %>%
    mutate(fraction=n/n.total) %>%
    filter(fraction > 0.15) %>%
    group_by(year) %>%
    summarize(vic_fraction=sum(fraction)),
    by="year") %>%
  ggplot(aes(x=year)) +
    ggtitle("Shooting Incidents, Black males age 18-44") +
    ggplot2::theme(
      legend.text=ggplot2::element_text(size=8),
      legend.box.margin = margin(6, 6, 6, 6),
      legend.position="bottom") +
    ggplot2::guides(color=ggplot2::guide_legend(ncol=1, title="")) +
    ylim(0, 1) +
    ylab("Fraction of Annual Shootings") +
    geom_point(aes(y=perp_fraction, color="Perpetrator (where known)")) +
    geom_line(aes(y=perp_fraction, color="Perpetrator (where known)")) +
    geom_point(aes(y=vic_fraction, color="Victim")) +
    geom_line(aes(y=vic_fraction, color="Victim")) +
    geom_hline(aes(yintercept=0.238, color="Entire Black Population (fraction of city)"), linetype=2)
```

::::

:::

# Conclusions - Demographic Conclusions

The impact of shootings (both as victim and perpetrator) falls wildly
disproportionately on individuals reported as Black males aged 18-44.

Any strategy or intervention must consider both these temporal and 
demographic factors or it will be mis-targeted or inefficient.

# A Caution

The demographic concentration of shootings should not be read as *causal*
but rather as descriptive and correlated. Demographics are correlated with
many other factors including socio-economic status, education, job opportunities,
geographic location, historical deprivation etc., which are more likely to
be actual causation.

# Error and Bias

Finally, we consider the many ways this analysis might be wrong or misleading.

# Error and Bias: Data Source Motivations

The underlying data comes from the NYPD. They are not a neutral party:
this data may be used to demonstrate their effectiveness, to argue for
more resources in certain areas, or to support theories of crime and
victimhood.

# Error and Bias: Data Source Avenues of Bias

Any of these would be a reason to bias the data collection and reporting:

* Skewing demographic data toward desirable patterns
* More subtly, controlling what is reported as a shooting incident and what is
  absent altogether

# Error and Bias: Analysis

Beyond the data, the analysis may be flawed, biased or misleading.

* The focus on proportional analysis masks the changing absolute numbers of
  incidents, which may be a better measure of "cost" in many dimensions
* There may be important subtleties in who reports data and how, unknown to
  the analyst. Demographic fields are subjective and vulnerable to misinterpretation.

That said, the patterns reported *seem* to be so strong as to resist most
small-scale biases or misinterpretations!